---
title: "ECO395M Final Project: Impact of Covid-19 on the flight delays and cancellation in California and Texas"
author: "Steven Kim and Shreekara Shastry"
date: ""
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(tidyverse)
library(knitr)
library(airportr)
library(lubridate)
library(randomForest)
library(rsample)
library(parallel)
library(MASS)
library(pscl)
library(purrr)
library(modelr)
library(pROC)
library(rpart)
library(rpart.plot)
library(vip)
library(gridExtra)
```

```{=latex}
\begin{titlepage}
\maketitle
\begin{abstract}
This is our abstract.
\end{abstract}
\thispagestyle{empty}
\end{titlepage}
\section{Introduction}
```

COVID-19 has affected our society in various ways. Amongst all of them, we decided to look at its effect on the 
flight delay and cancellations. 

```{=latex}
\section{Methods}
\subsection{Dataset}
```

We have used 2 separate datasets and combined them to form a data frame that we used in the further analysis. The first dataset is from The United States Department of Transportation's (DOT) Bureau of Transportation Statistics tracks the on-time performance of domestic flights operated by large air carriers. The data collected is from January to June 2020 and contains relevant flight information (on-time, delayed, canceled, diverted flights) from the Top 10 United States flight carriers for 11 million flights. The second dataset is from the New York Times[2] which contains the state-wise data on the daily number of new cases and deaths, the seven-day rolling average, and the seven-day rolling average per 100,000 residents. We merged these two datasets based on the date and state to create a new dataset that we used in all the models. This combined dataset has in total of 2745847 observations with data from 375 different airports.

```{=latex}
\subsection{Data Wrangling}
```

Data cleaning and preprocessing for the dataset was a four-step process.
1. Formatting the date field in both the individual datasets to match before performing a left join.
2. Merging the covid case dataset into the flight data based on date and state.
3. Factorizing the categorical variables from this combined dataset.  `MONTH`, `DAY_OF_MONTH`, `DAY_OF_WEEK`, `MKT_UNIQUE_CARRIER`, `TAIL_NUM`, `ORIGIN`, `ORIGIN_STATE_NM`, `DEST`, `DEST_STATE_NM`, `ARR_DEL15`, `CANCELLED`, `CANCELLATION_CODE`, these are the categorical variables in the dataset.
4. Removing the variables that are not used in the analysis to have a cleaner dataset.

```{r read-data}
airport = read.csv("Data/jantojun2020.csv")
```

Airline Carrier Code: AA: American Airlines AS: Alaska Airlines B6: JetBlue DL: Delta Air Lines F9: Frontier Airlines G4: Allegiant Air HA: Hawaiian Airlines NK: Spirit Airlines UA: United Airlines WN: Southwest Airlines

```{r, eval=FALSE}
airport_lookup("OGD", input_type = "IATA", output_type = "name")
airport_lookup("BRD", input_type = "IATA", output_type = "name")
```

```{r data-cleaning}
covid_2020 = read.csv("Data/us-states.csv") %>% mutate(year = substring(date,1,4)) %>% filter(year == 2020)
covid_2020 = covid_2020[order(as.Date(covid_2020$date, format="%Y/%m/%d")),]
covid_2020 = covid_2020 %>% dplyr::select(date, state, cases_avg_per_100k)
covid_2020$date = as.Date(covid_2020$date)

# changing the date format and the name of the origin state
airport$FL_DATE = airport$FL_DATE %>% as.Date(format = "%m/%d/%Y")

airport_covid = left_join(airport, covid_2020, by=c("FL_DATE" = "date", "ORIGIN_STATE_NM" = "state"))
airport_covid = airport_covid %>% mutate(cases_avg_per_100k = replace_na(cases_avg_per_100k, 0))

#dropping  irrelevant columns
cols = c("MONTH","DAY_OF_MONTH","DAY_OF_WEEK","FL_DATE", "MKT_UNIQUE_CARRIER", "MKT_CARRIER_FL_NUM","TAIL_NUM", "ORIGIN", "ORIGIN_STATE_NM", "DEST", "DEST_STATE_NM", "CRS_DEP_TIME", "CRS_ARR_TIME", "ARR_DELAY_NEW","ARR_DEL15", "CANCELLED", "CANCELLATION_CODE", "CRS_ELAPSED_TIME", "DISTANCE", "cases_avg_per_100k")
airport_covid = airport_covid[,cols]

#factorizing the categorical variables
cols1 = c("MONTH","DAY_OF_MONTH","DAY_OF_WEEK","MKT_UNIQUE_CARRIER", "TAIL_NUM", "ORIGIN", "ORIGIN_STATE_NM", "DEST", "DEST_STATE_NM","ARR_DEL15", "CANCELLED", "CANCELLATION_CODE")
airport_covid[cols1] <- lapply(airport_covid[cols1], as.factor)
```

```{r train-test-split}
#train/test split for California
airport_covid_California = airport_covid %>% filter(ORIGIN_STATE_NM == "California")
airport_covid_California$ORIGIN = droplevels(airport_covid_California$ORIGIN)
airport_covid_California_split = initial_split(airport_covid_California, 0.8)
airport_covid_California_train = training(airport_covid_California_split)
airport_covid_California_test = testing(airport_covid_California_split)

#train/test split for Texas
airport_covid_Texas = airport_covid %>% filter(ORIGIN_STATE_NM == "Texas")
airport_covid_Texas$ORIGIN = droplevels(airport_covid_Texas$ORIGIN)
airport_covid_Texas_split = initial_split(airport_covid_Texas, 0.8)
airport_covid_Texas_train = training(airport_covid_Texas_split)
airport_covid_Texas_test = testing(airport_covid_Texas_split)
```

```{r, scatterplot}
scatterCalifornia = ggplot(airport_covid_California) +
  geom_point(aes(x=cases_avg_per_100k, y=ARR_DELAY_NEW))
scatterTexas = ggplot(airport_covid_Texas) +
  geom_point(aes(x=cases_avg_per_100k, y=ARR_DELAY_NEW))
grid.arrange(scatterCalifornia, scatterTexas, ncol=2)
```

```{r delayrate-logit}
### Did Covid cases cause more delays?
## California
logitCalifornia_del = glm(ARR_DEL15 ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST_STATE_NM + DISTANCE + cases_avg_per_100k, data=airport_covid_California_train, na.action = na.omit, family="binomial")

logitCalifornia_del_pred = predict(logitCalifornia_del, airport_covid_California_test, type="response")  
# predicted probabilities
yhatCalifornia_del = ifelse(logitCalifornia_del_pred > 0.5, 1, 0)
confusion_out = table(y = airport_covid_California_test$ARR_DEL15, yhat = yhatCalifornia_del)
kable(confusion_out)

# out of sample accuracy for k=0.5
sum(diag(confusion_out))/sum(confusion_out)

#ROC curve not used
# rocobj <- plot.roc(airport_covid_California_test$CANCELLED, yhatCalifornia_del,
#                    main = "Confidence intervals", 
#                    percent=TRUE,
#                    ci = TRUE,                  # compute AUC (of AUC by default)
#                    print.auc = TRUE)           # print the AUC (will contain the CI)

## Texas
logitTexas_del = glm(ARR_DEL15 ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST_STATE_NM + DISTANCE + cases_avg_per_100k, data= airport_covid_Texas_train, na.action = na.omit, family="binomial")

logitTexas_del_pred = predict(logitTexas_del, airport_covid_Texas_test, type="response")  # predicted probabilities
yhatTexas_del = ifelse(logitTexas_del_pred > 0.5, 1, 0)
confusion_out = table(y = airport_covid_Texas_test$ARR_DEL15, yhat = yhatTexas_del)
kable(confusion_out)

# out of sample accuracy for k=0.5
sum(diag(confusion_out))/sum(confusion_out)

# ROC Curve not used
# rocobj <- plot.roc(airport_covid_Texas_test$CANCELLED, yhatTexas_del,
#                    main = "Confidence intervals", 
#                    percent=TRUE,
#                    ci = TRUE,                  # compute AUC (of AUC by default)
#                    print.auc = TRUE)           # print the AUC (will contain the CI)
```

```{r cancellationrate-logit}
## seeing the effect of covid on cancellations

# California
logitCalifornia = glm(CANCELLED ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST_STATE_NM + DISTANCE + cases_avg_per_100k, data= airport_covid_California_train, na.action = na.omit, family="binomial")

logitCalifornia_pred = predict(logitCalifornia, airport_covid_California_test, type="response")  # predicted probabilities
yhatCalifornia = ifelse(logitCalifornia_pred > 0.5, 1, 0)
confusion_out = table(y = airport_covid_California_test$CANCELLED, yhat = yhatCalifornia)
kable(confusion_out)

# out of sample accuracy for k=0.5
sum(diag(confusion_out))/sum(confusion_out)

# ROC Curve not used
# rocobj <- plot.roc(airport_covid_California_test$CANCELLED, yhatCalifornia,
#                    main = "Confidence intervals", 
#                    percent=TRUE,
#                    ci = TRUE,                  # compute AUC (of AUC by default)
#                    print.auc = TRUE)           # print the AUC (will contain the CI)

#Texas
logitTexas = glm(CANCELLED ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST_STATE_NM  + DISTANCE + cases_avg_per_100k, data= airport_covid_Texas_train, na.action = na.omit, family="binomial")

logitTexas_pred = predict(logitTexas, airport_covid_Texas_test, type="response")  # predicted probabilities
yhatTexas = ifelse(logitTexas_pred > 0.5, 1, 0)
confusion_out = table(y = airport_covid_Texas_test$CANCELLED, yhat = yhatTexas)
kable(confusion_out)

# out of sample accuracy for k=0.5
sum(diag(confusion_out))/sum(confusion_out)

# ROC Curve not used
# rocobj <- plot.roc(airport_covid_Texas_test$CANCELLED, yhatTexas,
#                    main = "Confidence intervals", 
#                    percent=TRUE,
#                    ci = TRUE,                  # compute AUC (of AUC by default)
#                    print.auc = TRUE)           # print the AUC (will contain the CI)
```

```{r cancellation-tree}
#prunefunction - not used (didn't make it better)
# prune_1se = function(my_tree) {
#   out = as.data.frame(my_tree$cptable)
#   thresh = min(out$xerror + out$xstd)
#   cp_opt = max(out$CP[out$xerror <= thresh])
#   prune(my_tree, cp=cp_opt)
# }

## California
california.tree.cancel = rpart(CANCELLED ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST_STATE_NM + DISTANCE + cases_avg_per_100k, data=airport_covid_California_train, control = rpart.control(cp = 0.002, minsplit=30), na.action = na.omit)

rpart.plot(california.tree.cancel, type=4)
vip(california.tree.cancel)

# RMSE value for out of sample
modelr::rmse(california.tree.cancel, airport_covid_California_test)

texas.tree.cancel <- rpart(CANCELLED ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST_STATE_NM  + DISTANCE + cases_avg_per_100k, data = airport_covid_Texas_train, na.action = na.omit)

rpart.plot(texas.tree.cancel, type=4)
vip(texas.tree.cancel)

# RMSE value for out of sample
modelr::rmse(texas.tree.cancel, airport_covid_Texas_test)
```

```{r delay-LM}
# linear regression on delay time (including 0)

## California
lmCalifornia <- lm(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + DISTANCE + cases_avg_per_100k, data = airport_covid_California_train, na.action = na.omit)
summary(lmCalifornia)
lmCalifornia_pred = predict(lmCalifornia, data = airport_covid_California_test)
sqrt(mean((airport_covid_California_test$ARR_DELAY_NEW %>% na.omit - lmCalifornia_pred)^2))

## Texas
lmTexas <- lm(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + DISTANCE + cases_avg_per_100k, data = airport_covid_Texas_train, na.action = na.omit)
lmTexas_pred = predict(lmTexas, data = airport_covid_Texas_test)
sqrt(mean((airport_covid_Texas_test$ARR_DELAY_NEW %>% na.omit - lmTexas_pred)^2))
```

Beta coefficient = 0.1170830. The partial effect of cases_avg_per_100k on the airline delay.

```{r delay-ZIP}
#zero inflatied poisson regression

## California
zipCalifornia <- zeroinfl(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + DISTANCE + cases_avg_per_100k | cases_avg_per_100k, data = airport_covid_California_train, na.action = na.omit)
summary(zipCalifornia)
zipCalifornia_pred = predict(zipCalifornia, type="response", data = airport_covid_California_test)
sqrt(mean((airport_covid_California_test$ARR_DELAY_NEW %>% na.omit - zipCalifornia_pred)^2))

## Texas
zipTexas <- zeroinfl(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + DISTANCE + cases_avg_per_100k | cases_avg_per_100k, data = airport_covid_Texas_train, na.action = na.omit)
summary(zipTexas)
zipTexas_pred = predict(zipTexas, type="response", data = airport_covid_Texas_test)
sqrt(mean((airport_covid_Texas_test$ARR_DELAY_NEW %>% na.omit - zipCalifornia_pred)^2))
```

For the zero-inflated poisson model, the first process generates zeros and the second process is governed by a Poisson distribution that generates counts, some of which may be zero. In this model building, the assumption is that the covid cases would generate the non-zero counts.

```{r delay-tree}
california.tree = rpart(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST_STATE_NM  + DISTANCE + cases_avg_per_100k, data=airport_covid_California_train, control = rpart.control(cp = 0.001, minsplit=30))

rpart.plot(california.tree, type=4)
vip(california.tree)

# RMSE value for out of sample
modelr::rmse(california.tree, airport_covid_California_test)

## Texas
texas.tree = rpart(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST_STATE_NM  + DISTANCE + cases_avg_per_100k, data=airport_covid_Texas_train, control = rpart.control(cp = 0.001, minsplit=30))

rpart.plot(texas.tree, type=4)
vip(texas.tree)

# RMSE value for out of sample
modelr::rmse(texas.tree, airport_covid_Texas_test)
```

```{r delaytime-ifdelayed-train-test-split}
#seeing the effect on delayed time given it's delayed for California
airport_covid_delayed = airport_covid %>% filter(ARR_DEL15 == 1)
airport_covid_delayed_California = airport_covid_delayed %>% filter(ORIGIN_STATE_NM == "California")
airport_covid_delayed_California$ORIGIN = droplevels(airport_covid_delayed_California$ORIGIN)
airport_covid_delayed_California_split = initial_split(airport_covid_delayed_California, 0.8)
airport_covid_delayed_California_train = training(airport_covid_delayed_California_split)
airport_covid_delayed_California_test = testing(airport_covid_delayed_California_split)

airport_covid_delayed_Texas = airport_covid %>% filter(ARR_DEL15 == 1) %>% filter(ORIGIN_STATE_NM == "Texas")
airport_covid_Texas$ORIGIN = droplevels(airport_covid_Texas$ORIGIN)
airport_covid_delayed_Texas_split = initial_split(airport_covid_delayed_Texas, 0.8)
airport_covid_delayed_Texas_train = training(airport_covid_delayed_Texas_split)
airport_covid_delayed_Texas_test = testing(airport_covid_delayed_Texas_split)
```

```{r rf-delaytime}
#random forest model
## California
rfCalifornia_delayed = randomForest(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + 
                                      MKT_UNIQUE_CARRIER + DISTANCE + cases_avg_per_100k,
                                    data=airport_covid_delayed_California_train, na.action=na.omit, 
                                    ntree = 500, importance = TRUE)
rmse(rfCalifornia_delayed, data=airport_covid_delayed_California_test)
varImpPlot(rfCalifornia_delayed, type=1)
partialPlot(rfCalifornia_delayed, airport_covid_delayed_California_test, 'cases_avg_per_100k')

## Texas
rfTexas_delayed = randomForest(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + 
                                  MKT_UNIQUE_CARRIER + DISTANCE + cases_avg_per_100k, 
                               data=airport_covid_delayed_Texas_train, na.action=na.omit, 
                               ntree = 500, importance = TRUE)
rmse(rfTexas_delayed, data=airport_covid_delayed_Texas_test)
varImpPlot(rfTexas_delayed, type=1)
partialPlot(rfTexas_delayed, airport_covid_delayed_Texas_test, 'cases_avg_per_100k')
```

Including `ORIGIN` only reduced the rMSE from 97.35954 to 97.20708. removing the variable for the entire country analysis, as there are too many levels.