---
title: "ECO395M Final Project"
author: "Steven Kim and Shreekara Shastry"
date: ""
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(tidyverse)
library(knitr)
library(airportr)
library(lubridate)
library(randomForest)
library(rsample)
library(parallel)
library(MASS)
```

```{r}
airport = read.csv("Data/jantojun2020.csv")
```

Let's look at the airlines in specific. In specific which airline has carried the most 
number of flights.

Airline Carrier Code: 
AA: American Airlines 
AS: Alaska Airlines 
B6: JetBlue 
DL: Delta Air Lines 
F9: Frontier Airlines 
G4: Allegiant Air 
HA: Hawaiian Airlines 
NK: Spirit Airlines 
UA: United Airlines 
WN: Southwest Airlines

```{r}
airport_sum = airport %>% group_by(MKT_UNIQUE_CARRIER) %>% summarise(total = n())

ggplot(airport_sum) +
  geom_col(aes(x=MKT_UNIQUE_CARRIER, y=total))
```

Let's analyse which airport has the most delays
```{r}
airport_del = na.omit(airport) %>% group_by(DEST) %>% summarise(arr_delay = mean(ARR_DELAY)) %>% arrange(desc(arr_delay))

ggplot(head(airport_del)) +
  geom_col(aes(x=DEST, y=arr_delay))
```


```{r}
airport_lookup("PPG", input_type = "IATA", output_type = "name")
airport_lookup("BRD", input_type = "IATA", output_type = "name")

airport_del = airport_del %>% mutate(name = airport_lookup(DEST, input_type = "IATA", output_type = "name"))
```


```{r}
covid_2020 = read.csv("Data/us-states.csv") %>% mutate(year = substring(date,1,4)) %>% filter(year == 2020)
covid_2020 = covid_2020[order(as.Date(covid_2020$date, format="%Y/%m/%d")),]
covid_2020 = covid_2020 %>% dplyr::select(date, state, cases_avg_per_100k)
covid_2020$date = as.Date(covid_2020$date)

# changing the date format and the name of the origin state
airport$FL_DATE = airport$FL_DATE %>% as.Date(format = "%m/%d/%Y")

airport_covid = left_join(airport, covid_2020, by=c("FL_DATE" = "date", "ORIGIN_STATE_NM" = "state"))
airport_covid = airport_covid %>% mutate(cases_avg_per_100k = replace_na(cases_avg_per_100k, 0))
```

Random Variables
```{r}
airport_covid_split = initial_split(airport_covid, 0.8)
airport_covid_train = training(airport_covid_split)
airport_covid_test = testing(airport_covid_split)

# starts = rep(100,40)
# fx = function() randomForest(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST + DISTANCE,data=airport_covid_train, na.action=na.omit)
# 
# numCores = detectCores()
# results = mclapply(starts, fx, mc.cores = numCores)
rm(airport, covid_2020)
forest1 = randomForest(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST + DISTANCE,data=airport_covid_train, na.action=na.omit)
```


