---
title: "ECO395M Final Project"
author: "Steven Kim and Shreekara Shastry"
date: ""
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(tidyverse)
library(knitr)
library(airportr)
library(lubridate)
library(randomForest)
library(rsample)
library(parallel)
library(MASS)
library(pscl)
library(caret)
library(parallel)
library(foreach)
library(purrr)
library(modelr)
library(pROC)
library(rpart)
library(rpart.plot)
```

```{r}
airport = read.csv("Data/jantojun2020.csv")
```

Let's look at the airlines in specific. In specific which airline has carried the most 
number of flights.

Airline Carrier Code: 
AA: American Airlines 
AS: Alaska Airlines 
B6: JetBlue 
DL: Delta Air Lines 
F9: Frontier Airlines 
G4: Allegiant Air 
HA: Hawaiian Airlines 
NK: Spirit Airlines 
UA: United Airlines 
WN: Southwest Airlines

```{r}
#total numbers of flights by each carrier
airport_sum = airport %>% group_by(MKT_UNIQUE_CARRIER) %>% summarise(total = n())

ggplot(airport_sum) +
  geom_col(aes(x=MKT_UNIQUE_CARRIER, y=total))
```

Let's analyse which airport has the most delays
```{r}
#proportion of delayed flights by airlines
airline_whether_delay = airport %>% group_by(MKT_UNIQUE_CARRIER) %>% summarise(delayed = mean(ARR_DEL15, na.rm=TRUE))
ggplot(airline_whether_delay) +
  geom_col(aes(x=MKT_UNIQUE_CARRIER, y=delayed))
```


```{r}
#average delay time of the delayed (more than 15 mins) flights by airlines
airline_delay = airport %>% filter(ARR_DEL15 == 1) %>% group_by(MKT_UNIQUE_CARRIER) %>% summarise(delay = mean(ARR_DELAY_NEW)) %>% arrange(desc(delay))
ggplot(airline_delay) +
  geom_col(aes(x=MKT_UNIQUE_CARRIER, y=delay))
```

```{r}
# top airports with most delays
airport_whether_delay = airport %>% group_by(DEST) %>% summarise(delayed = mean(ARR_DEL15, na.rm=TRUE)) %>% arrange(desc(delayed))
ggplot(head(airport_whether_delay)) +
  geom_col(aes(x=DEST, y=delayed))
```

```{r}
ggplot(tail(airport_whether_delay)) +
  geom_col(aes(x=DEST, y=delayed))
```

```{r}
#average delay time of the delayed (more than 15 mins) flights by airports
airport_delay = airport %>% filter(ARR_DEL15 == 1) %>% group_by(DEST) %>% summarise(delay = mean(ARR_DELAY_NEW)) %>% arrange(desc(delay))
ggplot(head(airport_delay)) +
  geom_col(aes(x=DEST, y=delay))
```

```{r}
ggplot(tail(airport_delay)) +
  geom_col(aes(x=DEST, y=delay))
```




```{r}
airport_lookup("OGD", input_type = "IATA", output_type = "name")
airport_lookup("BRD", input_type = "IATA", output_type = "name")

#airport_del = airport_del %>% mutate(name = airport_lookup(DEST, input_type = "IATA", output_type = "name"))
```


```{r}
covid_2020 = read.csv("Data/us-states.csv") %>% mutate(year = substring(date,1,4)) %>% filter(year == 2020)
covid_2020 = covid_2020[order(as.Date(covid_2020$date, format="%Y/%m/%d")),]
covid_2020 = covid_2020 %>% dplyr::select(date, state, cases_avg_per_100k)
covid_2020$date = as.Date(covid_2020$date)

# changing the date format and the name of the origin state
airport$FL_DATE = airport$FL_DATE %>% as.Date(format = "%m/%d/%Y")

airport_covid = left_join(airport, covid_2020, by=c("FL_DATE" = "date", "ORIGIN_STATE_NM" = "state"))
airport_covid = airport_covid %>% mutate(cases_avg_per_100k = replace_na(cases_avg_per_100k, 0))
cols = c("MONTH","DAY_OF_MONTH","DAY_OF_WEEK","FL_DATE", "MKT_UNIQUE_CARRIER", "MKT_CARRIER_FL_NUM","TAIL_NUM", "ORIGIN", "ORIGIN_STATE_NM", "DEST", "DEST_STATE_NM", "CRS_DEP_TIME", "CRS_ARR_TIME", "ARR_DELAY_NEW","ARR_DEL15", "CANCELLED", "CANCELLATION_CODE", "CRS_ELAPSED_TIME", "DISTANCE", "cases_avg_per_100k")
airport_covid = airport_covid[,cols]
cols1 = c("MONTH","DAY_OF_MONTH","DAY_OF_WEEK","MKT_UNIQUE_CARRIER", "TAIL_NUM", "ORIGIN", "ORIGIN_STATE_NM", "DEST", "DEST_STATE_NM","ARR_DEL15", "CANCELLED", "CANCELLATION_CODE")
airport_covid[cols1] <- lapply(airport_covid[cols1], as.factor)
airport_covid
```

Random Variables

```{r California}
#supervied models for California
airport_covid_California = airport_covid %>% filter(ORIGIN_STATE_NM == "California")
airport_covid_California$ORIGIN = droplevels(airport_covid_California$ORIGIN)
airport_covid_split = initial_split(airport_covid_California, 0.8)
airport_covid_train = training(airport_covid_split)
airport_covid_test = testing(airport_covid_split)
rfCalifornia = randomForest(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + DISTANCE + cases_avg_per_100k, data=airport_covid_train, na.action=na.omit, ntree = 50, mtry = 6, importance = TRUE)
```

```{r}
k_grid = seq(5,45,by=10)
cv_grid = foreach(k = k_grid, .combine='rbind') %do% {
  knn = knnreg(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST + DISTANCE + cases_avg_per_100k, airport_covid_train, k=k)
  rms = rmse(knn, airport_covid_test)
  c(k=k, err=rms)
} %>% as.data.frame

cv_grid_final = cv_grid %>% filter(err == min(cv_grid$err))
rownames(cv_grid_final) = c("KNN Model")
colnames(cv_grid_final) = c("k", "rMSE")
```

```{r}
#zero inflatied poisson regression for California
zipCalifornia <- zeroinfl(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST + DISTANCE + cases_avg_per_100k | cases_avg_per_100k, data = airport_covid_train)
summary(zipCalifornia)
```

```{r}
# seeing the effect of covid on cancellations
logitCalifornia = glm(CANCELLED ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST + DISTANCE + cases_avg_per_100k, data= airport_covid_train, na.action = na.omit, family="binomial")

logitCalifornia_pred = predict(logitCalifornia, airport_covid_test, type="response")  # predicted probabilities
yhatCalifornia = ifelse(logitCalifornia_pred > 0.5, 1, 0)
confusion_out = table(y = airport_covid_test$CANCELLED, yhat = yhatCalifornia)
kable(confusion_out)

# out of sample accuracy for k=0.5
sum(diag(confusion_out))/sum(confusion_out)

rocobj <- plot.roc(airport_covid_test$CANCELLED, yhatCalifornia,
                   main = "Confidence intervals", 
                   percent=TRUE,
                   ci = TRUE,                  # compute AUC (of AUC by default)
                   print.auc = TRUE)           # print the AUC (will contain the CI)
```
```{r}
california.tree = rpart(ARR_DELAY_NEW ~ MONTH + DAY_OF_WEEK + MKT_UNIQUE_CARRIER + ORIGIN + DEST + DISTANCE + cases_avg_per_100k, data=airport_covid_train, control = rpart.control(cp = 0.002, minsplit=30))

prune_1se = function(my_tree) {
  out = as.data.frame(my_tree$cptable)
  thresh = min(out$xerror + out$xstd)
  cp_opt = max(out$CP[out$xerror <= thresh])
  prune(my_tree, cp=cp_opt)
}

california.prune = prune_1se(california.tree)

# RMSE value for out of sample
modelr::rmse(california.prune, airport_covid_test)
```






