---
title: "ECO395M: Exercise 1"
author: "Steven Kim and Shreekara Shastry"
date: ""
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
library(mosaic)
library(ggmap)
library(airportr)
library(ggplot2)
library(rsample)
library(caret)
library(modelr)
library(parallel)
library(foreach)
library(ggrepel)
library(gifski)
library(gganimate)
```

## 1) Data visualization: flights at ABIA

```{r, echo = FALSE}
airport = read_csv('ABIA.csv')

all_airports_origin = unique(airport$Origin)
all_airports_dest = unique(airport$Origin)

all_unique_airports = unique(append(all_airports_origin, all_airports_dest))

# airpot_from_aus = airport %>%  filter(Origin == 'AUS')
airport_to_aus = airport %>% filter(Dest == 'AUS')

airports_by_month = airport_to_aus %>% group_by(Origin, Month) %>% summarize(count = n())
airports_by_month = airports_by_month %>% mutate(Latitude = airport_location(Origin)$Latitude,
                             Longitude = airport_location(Origin)$Longitude)
airports_by_month = airports_by_month %>% mutate(Month = month.name[Month])

map = get_map("usa", zoom=4)

map_month <- ggmap(map) + 
  geom_point(aes(x = Longitude, y = Latitude, size = sqrt(count)), data = airports_by_month, alpha = .4, color = "#ff8aab") + 
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  labs(x = '', y = '', title = "Origins of the Flights to AUS by Month",
       subtitle = '{closest_state}') +
  scale_size_area(name   = "Numbers of Flights", 
                  breaks = sqrt(c(1, 5, 10, 50, 100, 500)), labels = c(1, 5, 10, 50, 100, 500)) +
  geom_label_repel(data = airports_by_month %>% group_by(Origin, Latitude, Longitude) %>% summarize(count = sum(count)) %>% arrange(desc(count)) %>% head(5), label.padding = 0.1, box.padding = 0.1, label.size = 0.1, nudge_x = 0.002, nudge_y = +1.4, aes(x = Longitude, y = Latitude, label = Origin)) + 
  transition_states(Month, state_length = 5, transition_length = 3)
map_month
```

```{r, eval=FALSE, include=FALSE}
airports_by_day = airport_to_aus %>% group_by(Origin, DayOfWeek) %>% summarize(count = n())
airports_by_day = airports_by_day %>% mutate(Latitude = airport_location(Origin)$Latitude,
                             Longitude = airport_location(Origin)$Longitude)
day.name = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
airports_by_day = airports_by_day %>% mutate(day = day.name[DayOfWeek]) %>% arrange(DayOfWeek)


map_day <- ggmap(map) + 
  geom_point(aes(x = Longitude, y = Latitude, size = sqrt(count)), data = airports_by_day, alpha = .4, color = "#ff8aab") + 
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  labs(x = '', y = '', title = "Origins of the Flights to AUS by the Day of Week",
       subtitle = '{closest_state}') +
  scale_size_area(name   = "Numbers of Flights", 
                  breaks = sqrt(c(1, 5, 10, 50, 100, 500)), labels = c(1, 5, 10, 50, 100, 500)) +
  geom_label_repel(data = airports_by_day %>% group_by(Origin, Latitude, Longitude) %>% summarize(count = sum(count)) %>% arrange(desc(count)) %>% head(5), label.padding = 0.1, box.padding = 0.1, label.size = 0.1, nudge_x = 0.002, nudge_y = +1.4, aes(x = Longitude, y = Latitude, label = Origin)) + 
  transition_states(day, state_length = 5, transition_length = 3)
map_day
```

## 2) Wrangling the Billboard Top 100
Part A: Table of the top 10 most popular songs since 1958
```{r, echo=FALSE}
billboard = read_csv('billboard.csv')

billboard_A = billboard %>% 
  filter(year >= 1958) %>%
  group_by(song_id, song, performer) %>%
  summarize(count = n(), .groups = 'drop') %>%
  arrange(desc(count)) %>% head(10)  %>% select(song, performer, count)
billboard_A
```

Part B: Musical Diversity in Billboard Top 100
```{r, echo = FALSE}
billboard3 = c()

for (i in 1959:2020){
  billboard2 = billboard %>%
    filter(year == i) %>%
    group_by(song_id)
    billboard3[i-1958] = nrow(distinct(billboard2, song_id))
}

years = seq(1959, 2020, by=1)

billboard_4 = data.frame(years = years, unique_songs = billboard3)

billboard_B = ggplot(billboard_4) + geom_line(aes(x=years, y=unique_songs)) +
  labs(x = '', y = '', title = "Musical diversity of the Billboard Top 100, 1959-2020",
       caption = "Musical divserity has been calculated by counting the number of unique songs that appeared in the Billboard Top 100 that year.")
billboard_B
```

Part C: 30 Ten-Week Hits or more
```{r, echo = FALSE}
billboard5 = billboard %>%
  group_by(song_id, performer) %>%
  summarize(count = n(), .groups='drop') %>%
  filter(count >= 10) %>%
  group_by(performer) %>%
  summarize(count = n()) %>%
  filter(count >= 30)

bilboardC = ggplot(billboard5) + 
  geom_col(aes(x=fct_reorder(performer, count), y=count), color = 'pink') +
  coord_flip() +
  labs(x = '', y = 'Numbers of Ten-Week Hits', title = "Artists with 30 Or More Ten-Week Hits since 1958",
       caption = "A Ten-Week Hit is achieved when a single song appears on the Billboard Top 100 for at least ten weeks.")
bilboardC
```


## 3) Wrangling the Olympics

part A

```{r, echo=FALSE, results='hide'}
olympics = read_csv('olympics_top20.csv')

olympics_female = olympics %>%
  filter(sex == 'F') %>%
  filter(sport == 'Athletics') %>%
  summarize(q95_temp = quantile(height, 0.95))
```

The 95th percentile of heights for female competitors across all Athletics events is `r as.numeric(olympics_female)`.

part B

```{r, echo = FALSE}
olympics2 = olympics %>%
  filter(sex == 'F') %>%
  group_by(event) %>%
  summarize(sd_height = sd(height, na.rm=TRUE)) %>%
  filter(sd_height != 'NA') %>%
  arrange(desc(sd_height))
```

`r olympics2$event[1]` has had the greatest variability in competitor's heights across the entire history of the Olympics.

Part C

```{r, echo = FALSE}
olympics3 = olympics %>%
  filter(sport == 'Swimming') %>%
  group_by(year) %>%
  summarize(mean_age = mean(age))

olympics3$sex = "all"

olympics4 = olympics %>%
  filter(sport == 'Swimming') %>%
  group_by(year, sex) %>%
  summarize(mean_age = mean(age)) %>% rbind(olympics3) %>% arrange(year)

olympicsC = ggplot(olympics4) +
  geom_line(aes(x=year, y=mean_age, color = sex)) +
  labs(x = "", y = "", title = "Average Age of Olympic Swimmers Over Time",
       caption = "The average age of Olympic Swimmers deceased until 1950, stayed until 1975, and has been increasing since.
       Male swimmers tend to be older than female swimmers.")
olympicsC
```

## 4) K-nearest neighbors
### 350
```{r, echo=FALSE}
cars = read_csv('sclass.csv')

cars_350 = cars %>% filter(trim == '350')

# test train split
cars_350_split =  initial_split(cars_350, prop=0.9)
cars_350_train = training(cars_350_split)
cars_350_test  = testing(cars_350_split)

k_grid = seq(2,200, by=1)

cv_grid_350 = foreach(k = k_grid, .combine='rbind') %do% {
  knn = knnreg(price ~ mileage, data=cars_350_train, k=k)
  rms = rmse(knn, cars_350_test)
  c(k=k, err=rms)
} %>% as.data.frame

ggplot(cv_grid_350) + 
  geom_point(aes(x=k, y=err)) + 
  labs(y = 'RMSE') +
  scale_x_log10() +
  labs(title = "RMSE Plot for 350")

cv_grid_350_final = cv_grid_350 %>% filter(err == min(cv_grid_350$err))
```

The optimal value of k in this case is `r cv_grid_350_final$k`.

```{r, echo = FALSE}
knn_350 = knnreg(price ~ mileage, data=cars_350_train, k=cv_grid_350_final$k)

cars_350_test = cars_350_test %>%
  mutate(price_pred = predict(knn, cars_350_test))

plot_350_pred = ggplot(data = cars_350_test) + 
  geom_point(mapping = aes(x = mileage, y = price), alpha=0.2) +
  geom_line(aes(x = mileage, y = price_pred), color='red') +
  labs(title = "Plot of the Fitted Model for 350")
plot_350_pred
```

### 63 AMG
```{r, echo=FALSE}
cars_63AMG = cars %>% filter(trim == '63 AMG')

# test train split
cars_63AMG_split =  initial_split(cars_63AMG, prop=0.9)
cars_63AMG_train = training(cars_63AMG_split)
cars_63AMG_test  = testing(cars_63AMG_split)

cv_grid_63AMG = foreach(k = k_grid, .combine='rbind') %do% {
  knn = knnreg(price ~ mileage, data=cars_63AMG_train, k=k)
  rms = rmse(knn, cars_63AMG_test)
  c(k=k, err=rms)
} %>% as.data.frame

ggplot(cv_grid_63AMG) + 
  geom_point(aes(x=k, y=err)) + 
  labs(y = 'RMSE') +
  scale_x_log10() +
  labs(title = "RMSE Plot for 63AMG")

cv_grid_63AMG_final = cv_grid_63AMG %>% filter(err == min(cv_grid_63AMG$err))
```

The optimal value of k in this case is `r cv_grid_63AMG_final$k`.

```{r, echo = FALSE}
knn = knnreg(price ~ mileage, data=cars_63AMG_train, k=cv_grid_63AMG_final$k)

cars_63AMG_test = cars_63AMG_test %>%
  mutate(price_pred = predict(knn, cars_63AMG_test))

plot_63AMG_pred = ggplot(data = cars_63AMG_test) + 
  geom_point(mapping = aes(x = mileage, y = price), alpha=0.2) +
  geom_line(aes(x = mileage, y = price_pred), color='red') +
  labs(title = "Plot of the Fitted Model for 63AMG")
plot_63AMG_pred
```

63 AMG yields a larger optimal value of k. I suspect that it might be because there are a lot of 63AMGs with very low mileage.
